---
title: "funTDA — Part A: functional network baseline (H3N2)"
subtitle: "Persistence landscapes, FPCA, and functional permutation tests over the 17 GRNs"
author: "Victoria Arroniz"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 7
    fig-height: 5
  pdf:
    toc: true
    toc-depth: 3
    number-sections: false
    fig-width: 6.5
    fig-height: 4.5
    geometry: margin=1in
    pdf-engine: xelatex
engine: knitr
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(fda)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(igraph)

# Baseline reference values (for comparison only, NOT shown as a result:
# every number reported in the body of the document is recomputed from the
# CSVs committed under output/PH and data/GRN_adjacency_matrices).
ref <- list(
  pc1_pct = 63.5, pc2_pct = 20.0,
  test1_p = 0.0011, test1_Tobs = 6.88, test1_crit95 = 3.72,
  symp_mean = 0.496, symp_sd = 0.290, symp_frac05 = 5.6,
  asymp_mean = 0.481, asymp_sd = 0.291, asymp_frac05 = 5.7
)

# Unified color palette for all figures in the document
# Blue for the reference group, red for the contrast. These are the light pair
# of the make_figs.py palette (COL_BLUE_LIGHT / COL_RED_LIGHT), not the base R
# names "blue" and "red", which are pure primaries and read as harsh next to
# the figures make_figs.py draws. Change them here, not at the point of use.
col_asymp <- "#1a5fa8"
col_symp  <- "#C0392B"

# The memoria lives outside this repository, so a standalone clone must still
# render cleanly: figures are exported to it only when the target directory is
# actually present (checked at each call site). Three levels up, not two --
# this file sits one directory deeper than results_partB.qmd.
#
# These five figures used to be copied by hand out of
# results_partA_files/figure-pdf/, which only exists after a --to pdf render.
# Nothing recorded that step, so the memoria silently kept the 21 July versions
# through every later edit. Exporting them explicitly, as Part B already does,
# makes the export happen on any render, HTML included.
fig_dir <- "../../../ResearchProject_UCD/figures"
```

## Data

This analysis starts from 17 gene regulatory networks (GRNs) built from the
transcriptomic response to an experimental H3N2 influenza infection (Higgins,
Wu & Carey). Of the 17 subjects, 8 remained **asymptomatic** and 9 developed
**symptoms**. Each network is a 250×250 adjacency matrix, weighted and
**undirected** (symmetric).

```{r check-adjacency}
adj_dir <- "../data/GRN_adjacency_matrices"
adj_files <- list.files(adj_dir, full.names = TRUE)

adj_check <- lapply(adj_files, function(f) as.matrix(read.csv(f, header = FALSE)))
dims_ok <- all(vapply(adj_check, function(m) all(dim(m) == c(250, 250)), logical(1)))
max_asym <- max(vapply(adj_check, function(m) max(abs(m - t(m))), numeric(1)))

stopifnot(
  "Expected 17 adjacency matrices" = length(adj_files) == 17,
  "All matrices must be 250x250" = dims_ok,
  "All matrices must be symmetric (undirected)" = max_asym == 0
)
```

::: {.callout-note collapse="true"}
## Structural check of the data

- Number of adjacency matrices found: **`r length(adj_files)`** (8 asymptomatic + 9 symptomatic).
- Dimension of all matrices: **250 × 250** — `r if (dims_ok) "OK" else "FAIL"`.
- Maximum asymmetry `max(|A - A^T|)` over the 17 matrices: **`r max_asym`** (0 = perfectly symmetric / undirected).
:::

Before turning to persistent homology, it is worth looking at the networks
"by eye". The figure below compares, for one asymptomatic and one symptomatic
subject, a 60-node subgraph keeping only the highest-weight edges (threshold:
85th percentile of edge weights, same as in `make_figs.py`):

```{r network-comparison}
#| fig-cap: "60-node subgraphs (edges > 85th weight percentile) of one asymptomatic subject (subject 2) and one symptomatic subject (subject 1). No obvious structure visually distinguishes the two clinical classes — this motivates turning to persistent homology."
#| fig-width: 10
#| fig-height: 5
build_thresholded_graph <- function(adj_path, n_nodes = 60, seed = 0) {
  adj <- as.matrix(read.csv(adj_path, header = FALSE))
  set.seed(seed)
  nodes <- sort(sample(seq_len(nrow(adj)), n_nodes))
  sub <- adj[nodes, nodes]
  thr <- quantile(sub[upper.tri(sub)], 0.85)
  adj_thr <- sub
  adj_thr[adj_thr <= thr] <- 0
  igraph::graph_from_adjacency_matrix(adj_thr, mode = "upper", weighted = TRUE, diag = FALSE)
}

gA <- build_thresholded_graph(file.path(adj_dir, "Asymptomaticadjmatrix2.csv"))
gS <- build_thresholded_graph(file.path(adj_dir, "Symptomaticadjmatrix1.csv"))

w_rng <- range(c(igraph::E(gA)$weight, igraph::E(gS)$weight))
edge_width <- function(g) 0.4 + 1.6 * (igraph::E(g)$weight - w_rng[1]) / diff(w_rng)

set.seed(1); layoutA <- igraph::layout_in_circle(gA)
set.seed(1); layoutS <- igraph::layout_in_circle(gS)

draw_network_comparison <- function() {
  par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
  plot(gA, layout = layoutA, vertex.size = 5, vertex.label = NA,
       vertex.color = col_asymp, vertex.frame.color = col_asymp,
       edge.width = edge_width(gA), edge.color = grDevices::adjustcolor(col_asymp, alpha.f = 0.4),
       main = "Asymptomatic (subject 2)")
  plot(gS, layout = layoutS, vertex.size = 5, vertex.label = NA,
       vertex.color = col_symp, vertex.frame.color = col_symp,
       edge.width = edge_width(gS), edge.color = grDevices::adjustcolor(col_symp, alpha.f = 0.4),
       main = "Symptomatic (subject 1)")
  par(mfrow = c(1, 1))
}
draw_network_comparison()

# Exported for the memoria. Base R graphics, so pdf()/dev.off() rather than
# ggsave(); the size matches the chunk options above so the layout in the
# memoria does not shift.
if (dir.exists(fig_dir)) {
  pdf(file.path(fig_dir, "network_comparison.pdf"), width = 10, height = 5)
  draw_network_comparison()
  invisible(dev.off())
}
```

No structural difference stands out between the two panels: density, node
degree, and overall graph appearance are comparable despite belonging to
different clinical classes. This visual indistinguishability is precisely what
motivates describing each network through a topological invariant (persistent
homology) rather than direct graph statistics.

Persistence diagrams for each network (previously computed in Python with
`giotto-tda`, see `funTDA.ipynb`) are precomputed in `output/PH/*.csv`, with
columns `Birth, Death, dimension` and no header.

```{r check-dims}
ph_dir <- "../output/PH"
ph_files_raw <- list.files(ph_dir, pattern = "_PH.csv$", full.names = TRUE)

all_dims <- unlist(lapply(ph_files_raw, function(f) {
  d <- read.csv(f, header = FALSE)
  unique(d[[3]])
}))
unique_dims <- sort(unique(all_dims))

stopifnot(
  "Persistence diagrams should only contain dimensions 0 and 1" =
    identical(unique_dims, c(0, 1))
)
```

Inspecting the `dimension` column of the 17 persistence CSVs confirms that the
only dimensions present are **`r paste(unique_dims, collapse = ", ")`** (H0
connected components and H1 loops). The baseline computes only H0 and H1 (the
giotto-tda default, `homology_dimensions=(0,1)`), following funTDA; the analysis
below therefore uses H1 landscapes (order `r`=1, first layer `KK`=1), replicating
`funTDA.R`. The H2 / voids question is taken up with the image data, not the
network baseline.

## Pipeline: from persistence diagrams to functional landscapes

```{r landscape-functions}
tseq <- seq(0, 1, length.out = 500)

# Pure-R reimplementation of TDA::landscape(..., KK = 1): avoids the heavy
# TDA/GUDHI dependency. Formula (validated bit-for-bit against TDA::landscape,
# see appendix): lambda_1(t) = max_i max(0, min(t - b_i, d_i - t)) over the
# bars of the requested dimension.
r_landscape1 <- function(PH, dim = 1, tseq) {
  bd <- PH[PH$dimension == dim, c("Birth", "Death"), drop = FALSE]
  if (nrow(bd) == 0) return(rep(0, length(tseq)))
  b <- bd$Birth
  d <- bd$Death
  sapply(tseq, function(t) max(0, max(pmin(t - b, d - t))))
}

get_number <- function(fname) as.numeric(sub(".*adjmatrix([0-9]+)_PH.*", "\\1", fname))
ph_files <- ph_files_raw[order(sapply(ph_files_raw, get_number))]

DataSymp <- c()
DataAsymp <- c()
for (f in ph_files) {
  PH <- read.csv(f, header = FALSE)
  colnames(PH) <- c("Birth", "Death", "dimension")
  x1 <- r_landscape1(PH, dim = 1, tseq = tseq)

  fname <- basename(f)
  if (grepl("^Asymptomatic", fname)) {
    DataAsymp <- cbind(x1, DataAsymp)
  } else if (grepl("^Symptomatic", fname)) {
    DataSymp <- cbind(x1, DataSymp)
  }
}
DataSymp  <- t(DataSymp)
DataAsymp <- t(DataAsymp)
DataBoth  <- rbind(DataAsymp, DataSymp)

stopifnot(
  "Expected 8 asymptomatic subjects" = nrow(DataAsymp) == 8,
  "Expected 9 symptomatic subjects"  = nrow(DataSymp) == 9
)

# B-spline smoothing identical to funTDA.R: order-2 basis (piecewise linear),
# nbasis = Q - 1 = 499 knots over [0, 1].
create_fd <- function(Data) {
  x <- seq(0, 1, length.out = ncol(Data))
  basis <- create.bspline.basis(c(0, 1), nbasis = ncol(Data) - 1, norder = 2)
  fdobj <- smooth.basis(x, t(Data), basis)
  fdobj$fd
}

fdSymp  <- create_fd(DataSymp)
fdAsymp <- create_fd(DataAsymp)
fdBoth  <- create_fd(DataBoth)
```

The `landscape(..., KK = 1)` reimplementation was validated by comparing it
bit-for-bit against `TDA::landscape` over the 17 subjects (maximum absolute
difference = 0; see reproducible appendix). Curves are smoothed with an
order-2 B-spline basis (`nbasis = 499`) over `tseq = seq(0, 1, length.out = 500)`,
identical to `funTDA.R`.

### Example: persistence diagram and landscape of a representative subject

To illustrate the step from "persistence diagram" to "functional landscape",
a single representative subject is used — Symptomatic 1, the same one used in
`make_figs.py` for `network_to_landscape.png` — and both representations are
shown side by side, using the same `r_landscape1()` defined above.

```{r diagram-landscape-example}
#| fig-cap: "Representative subject (Symptomatic 1): H0/H1 persistence diagram (left) and H1 landscape $\\lambda_1(t)$ (right)."
#| fig-width: 9
#| fig-height: 4.5
PH_s1 <- read.csv(file.path(ph_dir, "Symptomaticadjmatrix1_PH.csv"), header = FALSE)
colnames(PH_s1) <- c("Birth", "Death", "dimension")

diagram_plot <- ggplot(PH_s1, aes(x = Birth, y = Death, shape = factor(dimension))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 2, alpha = 0.7, color = col_symp) +
  scale_shape_manual(values = c("0" = 16, "1" = 17), name = "Dimension",
                      labels = c("H0", "H1")) +
  labs(x = "Birth", y = "Death", title = "Persistence diagram") +
  coord_equal() +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ls_s1 <- r_landscape1(PH_s1, dim = 1, tseq = tseq)
landscape_plot <- ggplot(data.frame(t = tseq, lambda = ls_s1), aes(x = t, y = lambda)) +
  geom_line(color = col_symp, linewidth = 0.9) +
  labs(x = "t", y = expression(lambda[1](t)), title = "Persistence landscape (H1)") +
  theme_classic(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

diagram_landscape_example <- diagram_plot + landscape_plot
diagram_landscape_example

# Exported for the memoria (ResearchProject_UCD/figures/diagram_landscape_example.pdf).
if (dir.exists(fig_dir)) {
  ggsave(file.path(fig_dir, "diagram_landscape_example.pdf"), diagram_landscape_example,
         width = 9, height = 4.5)
}
```

### Landscapes by group

```{r group-landscape-comparison}
#| fig-cap: "Individual H1 landscapes, one panel per group (same vertical scale). Asymptomatic subjects consistently occupy a band of higher peaks (~0.11–0.15), while symptomatic subjects are more variable and generally lower (~0.02–0.12) — a pattern consistent with the PC1 separation and with the permutation test (p = 0.0011)."
indiv_list <- list()
for (i in seq_len(nrow(DataAsymp))) {
  indiv_list[[length(indiv_list) + 1]] <- data.frame(
    t = tseq, lambda = DataAsymp[i, ], subject = paste0("A", i), Group = "Asymptomatic"
  )
}
for (i in seq_len(nrow(DataSymp))) {
  indiv_list[[length(indiv_list) + 1]] <- data.frame(
    t = tseq, lambda = DataSymp[i, ], subject = paste0("S", i), Group = "Symptomatic"
  )
}
indiv_df <- do.call(rbind, indiv_list)
y_max <- max(indiv_df$lambda) * 1.05

plot_group_landscapes <- function(df, group_label, col) {
  ggplot(df, aes(x = t, y = lambda, group = subject)) +
    geom_line(color = col, alpha = 0.6, linewidth = 0.6) +
    scale_y_continuous(limits = c(0, y_max)) +
    labs(x = "t", y = expression(lambda[1](t)), title = group_label) +
    theme_classic(base_size = 13) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
}

plot_asymp <- plot_group_landscapes(indiv_df[indiv_df$Group == "Asymptomatic", ], "Asymptomatic", col_asymp)
plot_symp  <- plot_group_landscapes(indiv_df[indiv_df$Group == "Symptomatic", ], "Symptomatic", col_symp)

group_landscape_comparison <- plot_asymp + plot_symp
group_landscape_comparison

# Exported for the memoria (ResearchProject_UCD/figures/group_landscape_comparison.pdf).
if (dir.exists(fig_dir)) {
  ggsave(file.path(fig_dir, "group_landscape_comparison.pdf"), group_landscape_comparison,
         width = 6.5, height = 4.5)
}
```

## Functional Principal Component Analysis (FPCA)

```{r fpca}
pca <- pca.fd(fdBoth, nharm = 2)

pc1_pct <- pca$varprop[1] * 100
pc2_pct <- pca$varprop[2] * 100
cum_pct <- pc1_pct + pc2_pct

m <- c("A1", "A2", "A3", "A4", "A5", "A6", "A7", "A8",
       "S1", "S2", "S3", "S4", "S5", "S6", "S7", "S8", "S9")
group <- c(rep("Asymptomatic", 8), rep("Symptomatic", 9))

pca_df <- data.frame(
  PC1 = pca$scores[, 1],
  PC2 = pca$scores[, 2],
  ID = m,
  Group = group
)

n_asymp_pos <- sum(pca_df$PC1[pca_df$Group == "Asymptomatic"] > 0)
n_symp_neg  <- sum(pca_df$PC1[pca_df$Group == "Symptomatic"] < 0)
```

```{r fpca-table}
#| tbl-cap: "Variance explained by the first two functional components"
knitr::kable(
  data.frame(
    Component = c("PC1", "PC2", "Cumulative (PC1+PC2)"),
    `% variance explained` = c(sprintf("%.2f%%", pc1_pct), sprintf("%.2f%%", pc2_pct), sprintf("%.2f%%", cum_pct)),
    check.names = FALSE
  ),
  align = "lr"
)
```

```{r fpca-plot}
#| fig-cap: "PC1 vs PC2 scores (FPCA over H1 landscapes), colored by group"
fpca_score_plot <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Group, label = ID)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_point(size = 4, alpha = 0.9) +
  ggrepel::geom_text_repel(size = 4, show.legend = FALSE, max.overlaps = Inf) +
  scale_color_manual(values = c("Asymptomatic" = col_asymp, "Symptomatic" = col_symp)) +
  labs(x = "PC1 Score", y = "PC2 Score", title = "") +
  coord_fixed() +
  theme_classic(base_size = 16) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
fpca_score_plot

# Exported for the memoria (ResearchProject_UCD/figures/fpca_scores.pdf).
if (dir.exists(fig_dir)) {
  ggsave(file.path(fig_dir, "fpca_scores.pdf"), fpca_score_plot,
         width = 6.5, height = 4.5)
}
```

::: {.callout-note collapse="true"}
## Full FPC scores (all 17 subjects)

```{r fpca-full-scores}
pca_full_df <- data.frame(
  ID = pca_df$ID,
  Group = pca_df$Group,
  PC1 = round(pca_df$PC1, 4),
  PC2 = round(pca_df$PC2, 4)
)
knitr::kable(pca_full_df, align = "llrr")
```
:::

::: {.callout-note collapse="true"}
## Acceptance criteria — FPCA (baseline vs. recomputed)

| Metric | Reference | Recomputed |
|---|---:|---:|
| PC1 | `r sprintf("%.1f%%", ref$pc1_pct)` | `r sprintf("%.2f%%", pc1_pct)` |
| PC2 | `r sprintf("%.1f%%", ref$pc2_pct)` | `r sprintf("%.2f%%", pc2_pct)` |
| Cumulative | `r sprintf("%.1f%%", ref$pc1_pct + ref$pc2_pct)` | `r sprintf("%.2f%%", cum_pct)` |
| Asymptomatic with PC1 > 0 | 8 / 8 | `r n_asymp_pos` / 8 |
| Symptomatic with PC1 < 0 | 8 / 9 | `r n_symp_neg` / 9 |

```{r fpca-asserts}
tol_fpca <- 0.1  # percentage points
stopifnot(
  "PC1 out of tolerance relative to baseline" = abs(pc1_pct - ref$pc1_pct) < tol_fpca,
  "PC2 out of tolerance relative to baseline" = abs(pc2_pct - ref$pc2_pct) < tol_fpca,
  "Expected all 8 asymptomatic subjects with PC1 > 0" = n_asymp_pos == 8,
  "Expected exactly 8 of 9 symptomatic subjects with PC1 < 0" = n_symp_neg == 8
)
```
:::

The first two functional components explain **`r sprintf("%.1f", cum_pct)`%**
of the variability among landscapes (PC1 $\approx$ `r sprintf("%.1f", pc1_pct)`%, PC2 $\approx$
`r sprintf("%.1f", pc2_pct)`%). In the PC1–PC2 plane, all 8 asymptomatic subjects
fall in the PC1 > 0 half-plane, while 8 of the 9 symptomatic subjects fall in
PC1 < 0: **one of the symptomatic subjects** (labeled here in a purely
positional way — the `DataAsymp`/`DataSymp` construction in the
`landscape-functions` chunk above prepends (`cbind(x1, Data)`) each new
network to those already accumulated, reversing the order within each group
relative to the file ID, so the "S1..S9" label should not be read as the
subject's real ID without separately verifying it. `funTDA.R` was later fixed
to build these in ascending file-ID order with labels derived from the real
ID; this document keeps the original prepend-based construction so the
numbers below continue to match the validated baseline) overlaps with the
asymptomatic region. PC1 separates the two groups reasonably well; PC2 shows
no clear separation.

## Functional permutation test: asymptomatic vs. symptomatic

To avoid relying on `tperm.fd` with `nperm = 10000` (~40 s per test, infeasible
for the 196 split-half tests), a vectorized version of the **same test** is used:
it evaluates each functional object on the same grid `tperm.fd` uses by default
(`argvals = seq(0, 1, length.out = 101)`) and permutes columns of a matrix
instead of re-evaluating `fd` on every permutation. It was validated to
reproduce `tperm.fd` **bit-for-bit** with the same seed (see appendix).

```{r vtperm-def}
rowVar <- function(M) {
  n <- ncol(M)
  mu <- rowMeans(M)
  rowSums((M - mu)^2) / (n - 1)
}

vtperm <- function(X, i1, i2, nperm = 10000) {
  x1 <- X[, i1, drop = FALSE]
  x2 <- X[, i2, drop = FALSE]
  n1 <- ncol(x1); n2 <- ncol(x2)
  Xmat <- cbind(x1, x2)
  n <- n1 + n2
  Ts <- function(M) {
    m1 <- rowMeans(M[, 1:n1, drop = FALSE])
    m2 <- rowMeans(M[, n1 + (1:n2), drop = FALSE])
    v1 <- rowVar(M[, 1:n1, drop = FALSE]) / n1
    v2 <- rowVar(M[, n1 + (1:n2), drop = FALSE]) / n2
    max(abs(m1 - m2) / sqrt(v1 + v2))
  }
  Tobs <- Ts(Xmat)
  Tn <- numeric(nperm)
  for (i in 1:nperm) Tn[i] <- Ts(Xmat[, sample(n)])
  list(pval = mean(Tobs < Tn), Tobs = Tobs, crit95 = unname(quantile(Tn, 0.95)))
}
```

```{r permtests}
argvals <- seq(0, 1, length.out = 101)
cache_file <- "../output/results_partA_permtest_cache.rds"

if (file.exists(cache_file)) {
  cached <- readRDS(cache_file)
} else {
  set.seed(123)

  # 1) Main test: Asymptomatic vs Symptomatic
  XFull <- cbind(eval.fd(argvals, fdAsymp), eval.fd(argvals, fdSymp))
  test1 <- vtperm(XFull, 1:8, 9:17, nperm = 10000)

  # 2) Symptomatic split-half: combn(1:9, 4) = 126 4-vs-5 partitions
  XSymp <- eval.fd(argvals, fdSymp)
  splitsS <- combn(1:9, 4)
  pvalsS <- numeric(ncol(splitsS))
  for (i in seq_len(ncol(splitsS))) {
    g1 <- splitsS[, i]; g2 <- setdiff(1:9, g1)
    pvalsS[i] <- vtperm(XSymp, g1, g2, nperm = 10000)$pval
  }

  # 3) Asymptomatic split-half: combn(1:8, 4) = 70 4-vs-4 partitions.
  # Because the complement of a 4-of-8 subset is also a 4-of-8 subset, every
  # unique 4-vs-4 partition appears twice among these 70 (once with each
  # half as group1) — 35 unique partitions, each tested twice. Not
  # collapsed, to keep reproducing the baseline. The symptomatic loop above
  # (combn(1:9, 4), 4-vs-5) does not have this issue: a 5-element complement
  # is never itself drawn as a 4-element subset, so all 126 are distinct.
  XAsymp <- eval.fd(argvals, fdAsymp)
  splitsA <- combn(1:8, 4)
  pvalsA <- numeric(ncol(splitsA))
  for (i in seq_len(ncol(splitsA))) {
    g1 <- splitsA[, i]; g2 <- setdiff(1:8, g1)
    pvalsA[i] <- vtperm(XAsymp, g1, g2, nperm = 10000)$pval
  }

  cached <- list(test1 = test1, pvalsS = pvalsS, pvalsA = pvalsA)
  saveRDS(cached, cache_file)
}

test1  <- cached$test1
pvalsS <- cached$pvalsS
pvalsA <- cached$pvalsA

stopifnot(
  "Expected 126 symptomatic splits (combn(9,4))" = length(pvalsS) == 126,
  "Expected 70 asymptomatic splits (combn(8,4))" = length(pvalsA) == 70
)
```

::: {.callout-important}
## Asymptomatic vs. Symptomatic test result

- **p-value** = `r sprintf("%.4f", test1$pval)` (reference: `r sprintf("%.4f", ref$test1_p)`)
- **Observed T_max** = `r sprintf("%.2f", test1$Tobs)` (reference: `r sprintf("%.2f", ref$test1_Tobs)`)
- **Critical value (95%)** = `r sprintf("%.2f", test1$crit95)` (reference: `r sprintf("%.2f", ref$test1_crit95)`)

```{r test1-asserts}
tol_p <- 0.002
tol_T <- 0.05
stopifnot(
  "Main test p-value out of tolerance" = abs(test1$pval - ref$test1_p) < tol_p,
  "Observed Tmax out of tolerance" = abs(test1$Tobs - ref$test1_Tobs) < tol_T,
  "95% critical value out of tolerance" = abs(test1$crit95 - ref$test1_crit95) < tol_T
)
```
:::

With `T_max` = `r sprintf("%.2f", test1$Tobs)` well above the 95% critical value
(`r sprintf("%.2f", test1$crit95)`) and a p-value of `r sprintf("%.4f", test1$pval)`,
the null hypothesis of functional equality between the mean H1 landscapes of
asymptomatic and symptomatic subjects is rejected: **network topologies differ
statistically significantly between the two groups.**

## Split-half validation (test calibration)

If the permutation test is well calibrated, randomly splitting **the same
group** into two halves and comparing them against each other should yield
p-values approximately uniformly distributed on [0, 1] — that is, equality
should not be rejected more often than the nominal level (5%). This is checked
exhaustively by considering **all** possible partitions of each group:

```{r splithalf-table}
#| tbl-cap: "Summary of the split-half validation by group"
splithalf_summary <- data.frame(
  Group = c("Symptomatic (126 splits, 4 vs 5)", "Asymptomatic (70 splits, 4 vs 4)"),
  `mean p-value` = sprintf("%.3f", c(mean(pvalsS), mean(pvalsA))),
  `sd(p-value)` = sprintf("%.3f", c(sd(pvalsS), sd(pvalsA))),
  `% splits with p < 0.05` = sprintf("%.1f%%", c(mean(pvalsS < 0.05), mean(pvalsA < 0.05)) * 100),
  check.names = FALSE
)
knitr::kable(splithalf_summary, align = "lrrr")
```

```{r splithalf-hist}
#| fig-cap: "Distribution of split-half p-values, by group (dashed line at p = 0.05)"
histS <- ggplot(data.frame(p = pvalsS), aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0, fill = col_symp, alpha = 0.6, color = "white") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "black") +
  labs(title = "Symptomatic (126 splits)", x = "p-value", y = "Frequency") +
  theme_classic(base_size = 13)

histA <- ggplot(data.frame(p = pvalsA), aes(x = p)) +
  geom_histogram(binwidth = 0.05, boundary = 0, fill = col_asymp, alpha = 0.6, color = "white") +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "black") +
  labs(title = "Asymptomatic (70 splits)", x = "p-value", y = "Frequency") +
  theme_classic(base_size = 13)

splithalf_hist_combined <- histA + histS
splithalf_hist_combined

# Exported for the memoria (ResearchProject_UCD/figures/splithalf_hist.pdf).
if (dir.exists(fig_dir)) {
  ggsave(file.path(fig_dir, "splithalf_hist.pdf"), splithalf_hist_combined,
         width = 6.5, height = 4.5)
}
```

::: {.callout-note collapse="true"}
## Acceptance criteria — split-half (baseline vs. recomputed)

| Metric | Reference | Recomputed |
|---|---:|---:|
| Symptomatic — mean p | `r sprintf("%.3f", ref$symp_mean)` | `r sprintf("%.3f", mean(pvalsS))` |
| Symptomatic — sd p | `r sprintf("%.3f", ref$symp_sd)` | `r sprintf("%.3f", sd(pvalsS))` |
| Symptomatic — % p < 0.05 | `r sprintf("%.1f%%", ref$symp_frac05)` | `r sprintf("%.1f%%", mean(pvalsS < 0.05) * 100)` |
| Asymptomatic — mean p | `r sprintf("%.3f", ref$asymp_mean)` | `r sprintf("%.3f", mean(pvalsA))` |
| Asymptomatic — sd p | `r sprintf("%.3f", ref$asymp_sd)` | `r sprintf("%.3f", sd(pvalsA))` |
| Asymptomatic — % p < 0.05 | `r sprintf("%.1f%%", ref$asymp_frac05)` | `r sprintf("%.1f%%", mean(pvalsA < 0.05) * 100)` |

These quantities are Monte Carlo estimates (`nperm = 10000` per split)
aggregated over dozens of partitions, so variation of this kind is expected
relative to the baseline even when reproducing the algorithm exactly — in
fact, the main test above (computed with the same seed, in the same order)
reproduces the reference p-value exactly, and `vtperm` was validated
bit-for-bit against `tperm.fd`. That is why the tolerance here is looser than
in the main test.

```{r splithalf-asserts}
tol_mean <- 0.02
tol_sd   <- 0.02
tol_frac <- 2.0  # percentage points
stopifnot(
  "Symptomatic mean p out of tolerance"  = abs(mean(pvalsS) - ref$symp_mean) < tol_mean,
  "Symptomatic sd p out of tolerance"    = abs(sd(pvalsS) - ref$symp_sd) < tol_sd,
  "Symptomatic frac<0.05 out of tolerance" = abs(mean(pvalsS < 0.05) * 100 - ref$symp_frac05) < tol_frac,
  "Asymptomatic mean p out of tolerance" = abs(mean(pvalsA) - ref$asymp_mean) < tol_mean,
  "Asymptomatic sd p out of tolerance"   = abs(sd(pvalsA) - ref$asymp_sd) < tol_sd,
  "Asymptomatic frac<0.05 out of tolerance" = abs(mean(pvalsA < 0.05) * 100 - ref$asymp_frac05) < tol_frac
)
```
:::

In both groups, the fraction of splits with p < 0.05 (`r sprintf("%.1f%%", mean(pvalsS < 0.05) * 100)`
symptomatic, `r sprintf("%.1f%%", mean(pvalsA < 0.05) * 100)` asymptomatic) is
consistent with the nominal 5% expected under H0, and mean p-values hover
around 0.5 — the test does **not** produce systematic false positives when
comparing sub-samples of the same group, which calibrates the significance
reported for the asymptomatic–symptomatic contrast in the previous section.

## Appendix

### Validation of the `landscape(..., KK = 1)` reimplementation

```{r landscape-validation}
#| eval: false
#| echo: true
library(TDA)
tseq <- seq(0, 1, length.out = 500)
maxdiff <- 0
for (f in ph_files) {
  PH <- read.csv(f, header = FALSE)
  colnames(PH) <- c("Birth", "Death", "dimension")
  PHm <- as.matrix(PH[c("dimension", "Birth", "Death")])
  ref_landscape  <- landscape(PHm, dimension = 1, KK = 1, tseq)
  mine_landscape <- r_landscape1(PH, dim = 1, tseq = tseq)
  maxdiff <- max(maxdiff, max(abs(ref_landscape - mine_landscape)))
}
maxdiff  # 0 across the 17 subjects
```

### Validation of `vtperm` against `tperm.fd`

```{r vtperm-validation}
#| eval: false
#| echo: true
set.seed(123)
ref_test  <- tperm.fd(fdAsymp, fdSymp, nperm = 500, q = 0.05, plotres = FALSE)
set.seed(123)
mine_test <- vtperm(cbind(eval.fd(argvals, fdAsymp), eval.fd(argvals, fdSymp)),
                     1:8, 9:17, nperm = 500)
isTRUE(all.equal(ref_test$pval, mine_test$pval)) # TRUE
isTRUE(all.equal(ref_test$Tobs, mine_test$Tobs)) # TRUE
```

### Full code (reproducible)

```{r all-code}
#| ref.label: !expr knitr::all_labels()
#| echo: true
#| eval: false
```

### Session information

```{r session-info}
#| echo: true
sessionInfo()
```
